---
title: "EDA and Factor Analysis"
author: "Alex Stransky"
date: "November 22, 2017"
output: html_notebook
---

## Loading Data and Packages

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(psych)
```

Let's load the data and convert the date column to a date format (could be necessary for plotting results later on).

```{r}
setwd("~/R/Betting Project")
bigspreaddata <- read.csv("bigspreaddata.csv")
bigspreaddata$GAME.DATE <- mdy(bigspreaddata$GAME.DATE)
```

We also want to create variables for the difference between the home and road team factors. This will likely be best for predicting the spread between home and road teams.

```{r}
spreaddiff <- bigspreaddata %>%
  mutate(
    dOeFG = 100*(R.OeFG - H.OeFG),
    dOTOV = R.OTOV - H.OTOV,
    dOREB = R.OREB - H.OREB,
    dOFT = 100*(R.OFT - H.OFT),
    dDeFG = 100*(R.DeFG - H.DeFG),
    dDTOV = R.DTOV - H.DTOV,
    dDREB = R.DREB - H.DREB,
    dFT = 100*(R.DFT - H.DFT)
  ) %>%
  mutate(
    dOeFGL5 = 100*(R.OeFGL5 - H.OeFGL5),
    dOTOVL5 = R.OTOVL5 - H.OTOVL5,
    dOREBL5 = R.OREBL5 - H.OREBL5,
    dOFTL5 = 100*(R.OFTL5 - H.OFTL5),
    dDeFGL5 = 100*(R.DeFGL5 - H.DeFGL5),
    dDTOVL5 = R.DTOVL5 - H.DTOVL5,
    dDREBL5 = R.DREBL5 - H.DREBL5,
    dFTL5 = 100*(R.DFTL5 - H.DFT)
  ) %>%
  mutate(
    dPIE = R.PIE - H.PIE
  ) %>%
  mutate(
    dPPG = R.PPG - H.PPG,
    dPPGA = R.PPGA - H.PPGA,
    dPPGL5 = R.PPGL5 - H.PPGL5,
    dPPGAL5 = R.PPGAL5 - H.PPGAL5
  ) %>%
  mutate(
    dORTG = R.ORTG - H.ORTG,
    dDRTG = R.DRTG - H.DRTG,
    dORTGL5 = R.ORTGL5 - H.ORTGL5,
    dDRTGL5 = R.DRTGL5 - H.DRTGL5
  ) %>%
  select(SPREAD, TOTAL, Diff, ActTotal, H.PTS, R.PTS, dOeFG:dPIE, dPPG:dDRTGL5)
```

## Exploring the data using psych

```{r}
describe(spreaddiff)
```

```{r}
headTail(spreaddiff)
```

```{r}
quickView(spreaddiff)
```

To look at patterns in the data, let's use pairs.panels

```{r}
pairs.panels(spreaddiff[,c(3, 7:10, 23)])
```

Seems like the effective field goal percentage statistic is the most significantly correlated with the point differential. Let's see how the rating statistics correlate with point differential.

```{r}
pairs.panels(spreaddiff[,c(3, 24:25, 28:29)])
```

Much better correlation values for all 4 rating deltas between teams. As we would expect, the correlation between the DefRtg stats and point differential is negative, as teams with higher DefRtg are worse defensively. When DefRtg is positive, that means the Road team has a worse defense and vice versa. Meanwhile, Diff is the spread from the Home team perspective, meaning a positive DefRtg should correlate to a more negative Diff.

Let's also see how the points per game and points per game allowed stats correlate with point difference.

```{r}
pairs.panels(spreaddiff[,c(3, 23:26)])
```

Also pretty solid correlations. The problem is, these stats are probably highly correlated with offensive and defensive rating stats.

```{r}
pairs.panels(spreaddiff[,23:30])
```

As expected, all of the rating stats are highly correlated with their logical points per game corrolaries (i.e. OffRtg is intimately connected with PPG since the amount of points scored by the offense is a key input in both).

I expect these results to be largely the same if we were to just use the raw home and road numbers without taking into account the difference between the two.

## Removing outliers

This reminds me that we might want to remove some of the outlier points. Every sport involves those games where one team is completely unprepared to play and gets blown out by 30+ points. For the purpose of this model, it might be beneficial to rule out those observations with extreme actual point differentials.

The standard deviation of the point diff was 13.50 above. It seems reasonable to exclude those points with extreme point differences, falling more than 2 standard deviations outside of the predicted Vegas spread.

```{r}
exclouts <- bigspreaddata %>%
  mutate(sprError = Diff - SPREAD) %>%
  arrange(desc(abs(sprError)))
exclouts %>%
  filter(abs(sprError) > 27) %>%
  select(HOME, ROAD, GAME.DATE, SPREAD, H.PTS, R.PTS, sprError) %>%
  arrange(desc(abs(sprError)))
```

Given the previously mentioned criteria of excluding points with spread errors greater than 27, we would exclude 111 observations. Let's delve further into the sprError statistic to get a better idea of how often and by how much the spread usually misses.

```{r}
describe(exclouts$sprError)
```

Given that the standard deviation of the stat is only 12, we could probably trim further values, but let's stick with the initial criteria.

```{r}
exclouts <- exclouts %>%
  filter(abs(sprError) <= 27)
# We'll also want to create a table like spreaddiff without the outliers
sdexclouts <- exclouts %>%
  mutate(
    dOeFG = 100*(R.OeFG - H.OeFG),
    dOTOV = R.OTOV - H.OTOV,
    dOREB = R.OREB - H.OREB,
    dOFT = 100*(R.OFT - H.OFT),
    dDeFG = 100*(R.DeFG - H.DeFG),
    dDTOV = R.DTOV - H.DTOV,
    dDREB = R.DREB - H.DREB,
    dFT = 100*(R.DFT - H.DFT)
  ) %>%
  mutate(
    dOeFGL5 = 100*(R.OeFGL5 - H.OeFGL5),
    dOTOVL5 = R.OTOVL5 - H.OTOVL5,
    dOREBL5 = R.OREBL5 - H.OREBL5,
    dOFTL5 = 100*(R.OFTL5 - H.OFTL5),
    dDeFGL5 = 100*(R.DeFGL5 - H.DeFGL5),
    dDTOVL5 = R.DTOVL5 - H.DTOVL5,
    dDREBL5 = R.DREBL5 - H.DREBL5,
    dFTL5 = 100*(R.DFTL5 - H.DFT)
  ) %>%
  mutate(
    dPIE = R.PIE - H.PIE
  ) %>%
  mutate(
    dPPG = R.PPG - H.PPG,
    dPPGA = R.PPGA - H.PPGA,
    dPPGL5 = R.PPGL5 - H.PPGL5,
    dPPGAL5 = R.PPGAL5 - H.PPGAL5
  ) %>%
  mutate(
    dORTG = R.ORTG - H.ORTG,
    dDRTG = R.DRTG - H.DRTG,
    dORTGL5 = R.ORTGL5 - H.ORTGL5,
    dDRTGL5 = R.DRTGL5 - H.DRTGL5
  ) %>%
  select(SPREAD, TOTAL, Diff, ActTotal, H.PTS, R.PTS, dOeFG:dPIE, dPPG:dDRTGL5)
```

## Principal Component Analysis

```{r}
sdexclouts.pca1 <- princomp(sdexclouts[,c(7:31)])
hptmod <- exclouts[,c(8:15, 24:25, 38:46, 52, 55, 59, 62, 64, 66, 69, 71)]
rptmod <- exclouts[,c(16:24, 30:37, 46:47, 53, 56, 58, 61, 65, 67:68, 70)]
exclouts.pca2 <- princomp(hptmod[,c(1:27)])
exclouts.pca3 <- princomp(rptmod[,c(1:27)])
```

```{r}
summary(sdexclouts.pca1)
```

```{r}
biplot(sdexclouts.pca1, scale = 0)
```

It appears from the biplot that the most significant factors for the first component are the PPGAL5 and the DRTGL5. The second component seems to be highly dependent on PPGL5 and ORTGL5. Based on this, it seems like the top two components are primarily weighted by the core offensive and defensive statistics.

It seems like the overall season averages for ORTG/DRTG and PPG/PPGA are relatively important as well, but less so than the Last 5 statistics. In a later portion of this project, I'll attempt to develop weights for predicting what a teams OeFG, OTOV, etc. will be in a given game and use each teams in game projections to predict the spread between opponents.

```{r}
#Scree plot
pr_var1 <- (sdexclouts.pca1$sdev)^2
prop_varex1 <- pr_var1 / sum(pr_var1)
plot(cumsum(prop_varex1),
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Diff Scree Plot",
     type = "b")
abline(h = 0.95, col = "red", lty = 2)
```

Greater than 95% of the variance is explained with 11 components, and it seems like there are diminishing returns beyond that point.

```{r}
summary(exclouts.pca2)
```

```{r}
biplot(exclouts.pca2, scale = 0)
```

The first two components are less decisively made up of a couple of factors. However, there is a common theme in terms of the grouping of the most significant principal components. One group primarily depends on the performance of the road team's defense and the other primarily depends on the performance of the home teams offense (makes sense).

Let's find the best cutoff point for the number of factors we'll use in factor analysis.

```{r}
#Scree plot
pr_var2 <- (exclouts.pca2$sdev)^2
prop_varex2 <- pr_var2 / sum(pr_var2)
plot(cumsum(prop_varex2),
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Home Points Scree Plot",
     type = "b")
abline(h = 0.95, col = "red", lty = 2)
```

Seems like there are still gains to be made up through the 12th component even though the 10th component passes the 95% threshold.

```{r}
summary(exclouts.pca3)
```

```{r}
biplot(exclouts.pca3, scale = 0)
```

```{r}
#Scree plot
pr_var3 <- (exclouts.pca3$sdev)^2
prop_varex3 <- pr_var3 / sum(pr_var3)
plot(cumsum(prop_varex3),
     xlab = "Principal Component",
     ylab = "Proportion of Variance Explained",
     main = "Road Points Scree Plot",
     type = "b")
abline(h = 0.95, col = "red", lty = 2)
```

Finally, we'll take the principal component scores for the observations across all three principal component analyses. These will be used to test the value of using them over the raw predictors in LASSO regression. They won't be as necessary for Random Forest and XGBoost, as one of the major advantages of tree-based models is their robustness when dealing with highly correlated predictors.

We'll generate scores for the first 5, first 10, and first 15 principal components for all 3 sets of data.

```{r}
dscores5 <- as.data.frame(cbind(sdexclouts.pca1$scores[,1:5], sdexclouts$Diff))
dscores10 <- as.data.frame(cbind(sdexclouts.pca1$scores[,1:10], sdexclouts$Diff))
dscores15 <- as.data.frame(cbind(sdexclouts.pca1$scores[,1:15], sdexclouts$Diff))
hscores5 <- as.data.frame(cbind(exclouts.pca2$scores[,1:5], exclouts$H.PTS))
hscores10 <- as.data.frame(cbind(exclouts.pca2$scores[,1:10], exclouts$H.PTS))
hscores15 <- as.data.frame(cbind(exclouts.pca2$scores[,1:15], exclouts$H.PTS))
rscores5 <- as.data.frame(cbind(exclouts.pca3$scores[,1:5], exclouts$R.PTS))
rscores10 <- as.data.frame(cbind(exclouts.pca3$scores[,1:10], exclouts$R.PTS))
rscores15 <- as.data.frame(cbind(exclouts.pca3$scores[,1:15], exclouts$R.PTS))
```

The final thing we need to do is separate the home and road predictor variables for the exclouts data frame.

```{r}
exclouts.H <- exclouts[,c(1, 3, 6:7, 8:15, 24:25, 38:46, 52, 55, 59, 62, 64, 66, 69, 71:74)]
exclouts.R <- exclouts[,c(1, 3, 6:7, 16:24, 30:37, 46:47, 53, 56, 58, 61, 65, 67:68, 70, 72:74)]
```

```{r}
ggplot(data = bigspreaddata) +
  geom_histogram(aes(x = SPREAD), binwidth = 1.0, fill = "#00bfc4", alpha = 1.00) +
  geom_histogram(aes(x = Diff), binwidth = 1.0, fill = "#f8766d", alpha = 0.80) +
  fte +
  ggtitle("Point Spread and Final Score Distribution") +
  xlab("Spread") + ylab("Count") +
  geom_vline(xintercept = mean(bigspreaddata$Diff), colour = "#535353", lty = 2, size = 1) +
  geom_hline(yintercept = 0, colour = "#535353", size = 1.2) +
  annotate("text", x = 7, y = 250, label = "Mean Diff. (-2.74)", color = "#535353", size = 3.5, fontface = "bold") +
  annotate("text", x = -19, y = 195, label = "Closing Spread", color = "#00bfc4", size = 3.5, fontface = "bold") +
  annotate("text", x = 19.5, y = 115, label = "Actual Score Diff.", color = "#f8766d", size = 3.5, fontface = "bold") +
  scale_x_continuous(limits = c(-50, 40), breaks = seq(-45, 45, 10))
```


